Revealing transient strain in geodetic data with Gaussian process regression

### 2017-06-20 tex/hines_hetland_2017_gji.tex
added more

\documentclass{gji}
\usepackage{timet}

\title[Geophys.\ J.\ Int.: \LaTeXe\ Guide for Authors]
  {Geophysical Journal International:
   \LaTeXe\ style guide for authors}
\author[B.L.N. Kennett]
  {B.L.N. Kennett$^1$\thanks{Pacific Region Office, GJI} \\
  $^1$ Research School of Earth Sciences, Australian National
    University, Canberra ACT \emph{0200}, Australia
  }
\date{Received 1998 December 18; in original form 1998 November 22}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{200}
\pubyear{1998}

%\def\LaTeX{L\kern-.36em\raise.3ex\hbox{{\small A}}\kern-.15em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
%\def\LATeX{L\kern-.36em\raise.3ex\hbox{{\Large A}}\kern-.15em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
% Authors with AMS fonts and mssymb.tex can comment out the following
% line to get the correct symbol for Geophysical Journal International.
\let\leqslant=\leq

@@

\begin{summary}
 This guide is for authors who are preparing papers for
 \textit{Geophysical Journal International} using the
\LaTeXe\ document preparation system and the GJI class file.
\end{summary}

\begin{keywords}
 \LaTeXe\ -- class files: \verb"gji.cls"\ -- sample text -- user guide.
\end{keywords}

\section{Introduction}

In addition to the standard submission of hardcopy from authors,
\textit{Geophysical Journal International} accepts machine-readable forms
of  papers in \LaTeX.
The layout design for \textit{Geophysical Journal International} has been
implemented as a \LaTeXe\ class file derived from the MN style file for
Monthly Notices of the Royal Astronomical Society. The GJI classfile is
based on the \verb"ARTICLE" style as discussed in the \LaTeX\ manual
\cite {la}. Commands which differ from the standard \LaTeX\ interface, or
which are provided in addition to the standard interface, are explained
in this guide. This guide is not a substitute for the \LaTeX\ manual
itself. Authors planning to submit their papers in \LaTeX\ are advised to
use \verb"gji.cls" as  early as possible in the creation of their
files.  This guide is modified from that produced by Woollatt et al (1994)
to describe the features of the MN style.

A very accessible guide to the features of \LaTeXe and the differences
from the earlier version is provided by Kopka \& Daly \shortcite{kd}.  This
reference provides in chapter 9 a summary of \LaTeX\ error messages and also
a full description of standard \LaTeX\ commands in Appendix F.

@@

@@

\begin{thebibliography}{}
  \bibitem[\protect\citename{Butcher }1992]{bu}
    Butcher J. 1992. \textit{Copy-editing: The Cambridge
    Handbook}, 3rd edn, Cambridge Univ. Press, Cambridge.
  \bibitem[\protect\citename{The Chicago Manual }%
    1982]{cm} \textit{The Chicago Manual of Style}, Univ.
    Chicago Press, Chicago, 1982.
  \bibitem[\protect\citename{Chao }1985]{ch}
    Chao, B. F., 1985. Normal mode study of the Earth's rigid
     body motions, \textit{Geophys. Res. Lett.}, \textbf{12}, 526-529.
  \bibitem[\protect\citename{Hinderer }1986]{hi}
    Hinderer, J., 1986. Resonance effects of the earth's fluid
    core in earth rotation, in \textit{Solved and Unsolved
    Problems}, pp. 277-296, ed. Cazenave A., Reidel,
    Dordrecht.
  \bibitem[\protect\citename{Kopka \& Daly}1995]{kd}
    Kopka H. \& Daly P.W., 1995, \textit{A guide to} \LaTeX2e,
    Addison--Wesley, New York
  \bibitem[\protect\citename{Lamport }1986]{la}
    Lamport L., 1986,  \LaTeX: \textit{A Document
    Preparation System}, Addison--Wesley, New York
  \bibitem[\protect\citename{Lindberg }1986]{bl}
    Lindberg, C., 1986.  Multiple taper harmonic analysis of
    terrestrial free oscillations, \textit{PhD thesis},
    University of California.
  \bibitem[\protect\citename{Maupin }1992]{ma}
    Maupin, V., 1992. Modelling of laterally trapped surface
    waves with application to Rayleigh waves in the Hawaiian
    swell, \textit{Geophys. J. Int.}. \textbf{110}, 553-570.
  \bibitem[\protect\citename{Rutherford \& Hawker }%
    1981]{rh}  Rutherford, S. R. \& Hawker, K. E., 1981,
    Consistent coupled mode theory of sound propagation for a
    class of non-separable problems,
   \textit{J. acoust. Soc. Am.}, \textbf{71}, 554-564
\end{thebibliography}

>>>>

\documentclass{gji}
\usepackage{timet}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{arydshln}
\usepackage{bm}
\usepackage{ulem}

\title[Transient strain in geodetic data]
  {Revealing transient strain in geodetic data with Gaussian process regression}
\author{T. T. Hines$^1$ and E. A. Hetland$^1$ \\
        $^1$ Department of Earth and Environmental Sciences, University of Michigan, Ann Arbor, MI, USA}
\date{Received XXX; in original form XXX}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{XXX}
\pubyear{XXX}

% Authors with AMS fonts and mssymb.tex can comment out the following
% line to get the correct symbol for Geophysical Journal International.
\let\leqslant=\leq

\newcommand*{\du}[1]{\uuline{#1}}

@@

\begin{summary}
TODO
\end{summary}

\begin{keywords}
 TODO -- TODO -- TODO -- TODO.
\end{keywords}

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates estimated from GNSS data have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it feasible to estimate transient crustal strain, which we consider to be any deviation from the secular strain rate.  Transient crustal strain can potentially illuminate geophysical signal that may not be immediately apparent from the GNSS displacement data. The ability to detect transient geophysical phenomena is also relevant for assessing seismic hazard because there are instances of major earthquakes being triggered by slow slip events \citep{Roeloffs2006} and postseismic deformation \citep{Freed2001}.

%Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant deformation gradients) over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined through a least squares fit to the observations. The assumption that deformation gradients are spatially uniform is not appropriate when subnetworks span too large of an area. To help overcome this deficiency, \citet{Shen1996,Shen2015} used an inverse distance weighting scheme, in which the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996,Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The type of basis functions and the number of degrees of freedom for a parameterization can be subjective. If there are too few degrees of freedom in the parameterization, then estimated strain rates will be biased and the uncertainties will be underestimated. On the other hand, if there are too many degrees of freedom, then there will not be any coherent features in the estimated strain rates. The methods described by \citet{Beavan2001}, \citet{Tape2009}, and \citet{Sandwell2016} also require the user to specify penalty parameters that control a similar trade-off between bias and variance in the solution. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and fault geometries are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

%The aforementioned studies are concerned with estimating secular strain rates. In recent years the Southern California Earthquake Center (SCEC) community has shown interest in developing methods for detecting transient strain. SCEC supported a transient detection exercise \citep{Lohman2013}, where several research groups tested their methods for detecting transient geophysical signals with a synthetic GNSS dataset. Among the methods tested were the Network Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion Filter (NIF) \citep{Segall1997}. The NSF uses a wavelet parameterization to  describe the spatial component of geophysical signal. The NIF, which is intended for imaging slow fault slip from geodetic data, uses the elastic dislocation Green's functions from \citet{Okada1992}. For the NSF and NIF, the time dependence of the geophysical signal is modeled as integrated Brownian motion. The method described in \citet{Holt2013} was also tested in the SCEC transient detection exercise, which calculates strain rates using a bi-cubic spatial parameterization of displacements between time epochs. \citet{Holt2013} defined a detection threshold based on the strain rate magnitude, and below we demonstrate that this is indeed an effective criterion for identifying geophysical signal. For the same reasons descibed above, the transient deformation and corresponding uncertainties estimated by these methods can be biased by the chosen spatial parameterization. It is then difficult to distinguish signal from noise with these methods, which limits their utility for transient detection.

%Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Kato1998}, and it has recently gained popularity in the machine learning community. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed, continuous, stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. Here we use GPR to infer transient strain resulting from slow slip events (SSEs) in Cascadia. Our results demonstrate that GPR can be an effective tool for detecting SSEs and revealing the spatio-temporal evolution of crustal strain during SSEs.

@@

%\bibliographystyle{apalike}
%\bibliography{mybib}

@@

%\begin{thebibliography}{}
%  \bibitem[\protect\citename{Butcher }1992]{bu}
%    Butcher J. 1992. \textit{Copy-editing: The Cambridge
%    Handbook}, 3rd edn, Cambridge Univ. Press, Cambridge.
%  \bibitem[\protect\citename{The Chicago Manual }%
%    1982]{cm} \textit{The Chicago Manual of Style}, Univ.
%    Chicago Press, Chicago, 1982.
%  \bibitem[\protect\citename{Chao }1985]{ch}
%    Chao, B. F., 1985. Normal mode study of the Earth's rigid
%     body motions, \textit{Geophys. Res. Lett.}, \textbf{12}, 526-529.
%  \bibitem[\protect\citename{Hinderer }1986]{hi}
%    Hinderer, J., 1986. Resonance effects of the earth's fluid
%    core in earth rotation, in \textit{Solved and Unsolved
%    Problems}, pp. 277-296, ed. Cazenave A., Reidel,
%    Dordrecht.
%  \bibitem[\protect\citename{Kopka \& Daly}1995]{kd}
%    Kopka H. \& Daly P.W., 1995, \textit{A guide to} \LaTeX2e,
%    Addison--Wesley, New York
%  \bibitem[\protect\citename{Lamport }1986]{la}
%    Lamport L., 1986,  \LaTeX: \textit{A Document
%    Preparation System}, Addison--Wesley, New York
%  \bibitem[\protect\citename{Lindberg }1986]{bl}
%    Lindberg, C., 1986.  Multiple taper harmonic analysis of
%    terrestrial free oscillations, \textit{PhD thesis},
%    University of California.
%  \bibitem[\protect\citename{Maupin }1992]{ma}
%    Maupin, V., 1992. Modelling of laterally trapped surface
%    waves with application to Rayleigh waves in the Hawaiian
%    swell, \textit{Geophys. J. Int.}. \textbf{110}, 553-570.
%  \bibitem[\protect\citename{Rutherford \& Hawker }%
%    1981]{rh}  Rutherford, S. R. \& Hawker, K. E., 1981,
%    Consistent coupled mode theory of sound propagation for a
%    class of non-separable problems,
%   \textit{J. acoust. Soc. Am.}, \textbf{71}, 554-564
%\end{thebibliography}
